FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2024-12-16 manuscript.Rmd
some more thoughts

There have been notable efforts to improve theories' findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [@grayHowMapTheory2017].
Similarly, Borsboom and colleagues seek to establish a database of psychological theories [REF BORSBOOM].
Post-hoc curation does not address the root cause of the lack of Findability, however.
Ideally, Findability would be addressed ante-hoc, through modular publishing and documentation with rich metadata.

@@

The Accessibility criteria serve to *regulate* access, not to maximize it.
These principles apply to theory with minor changes, with the caveat that there might be less of a need to restrict access to theory than there is for (human subjects) data.
Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
This can be achieved, for example, by hosting theory in a Git remote repository, and archiving that repository on Zenodo for long-term storage.
The resulting resource will then have an identifier (DOI) which allows the theory to be accessed using a standardized communications protocol (download via `https` or `git`).
Secondly (A2), theory metadata should be accessible, even when the theory is no longer available,
which is also achieved via long-term storage (e.g., on Zenodo).
Git remote repositories allow for access control,
and Zenodo allows for access control of individual files/resources.

@@

A third impediment arises when theories have a "dependency on the author" (DOA).
DOA occurs when a theory cannot be understood by independent scholars,
thus requiring the original author for interpretation and clarification.
We have heard DOA referred to apocryphally as the "ask Leon" phenomenon,
as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
DOA relates to the discourse on "Great Man Theorizing" [@guestWhatMakesGood2024] because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
which violates checks and balances of scientific research.
DOA also renders theories immune to refutation,
because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target  [@szollosiArrestedTheoryDevelopment2021].
The fact that DOA is inherently problematic is illustrated by cases where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
This demonstrates that original authors are not the ultimate authority on their theories.
DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.

@@

Interoperability pertains to the property of information artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
The original interoperability principles can be rephrased somewhat to apply to theory.
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or multi-interpretable bitmap image falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated as 
as a type of data that is human- and machine-readable with as few interpretative steps as possible,
as previously recommended [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some axiomatic system.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very interoperable,
the same image represented in the DOT language for representing graphs does meet this ideal.

@@

If take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
Applied to FAIR theory, reusability requires that  theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1), detailed provenance (R1.2), and (meta)data which meets domain-relevant community standards (R1.3).
As we will argue below, the most appropriate license for theory reuse is likely to be CC0 (no rights reserved),
although this should be combined with a culture of comprehensive (theory) citation to meet other open science requirements [REF TOP guidelines].
Criterion R1.2 is met by version control with Git and archival on Zenodo.
Domain-relevant community standards, to a large extent, remain to be established - and this paper is the first step towards further work in that area.

>>>>

There have been notable efforts to improve theories' findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [@grayHowMapTheory2017].
Similarly, Borsboom and colleagues seek to establish a database of psychological theories [REF BORSBOOM].
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability, however.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.

@@

The Accessibility criteria serve to *regulate* access, not to maximize it.
These principles apply to theory with minor changes, with the caveat that there might be less of a need to restrict access to theory than there is for (human subjects) data.
Firstly, theory and its associated metadata should be accessible by their identifier using a standardized communications protocol (A1).
This can be achieved, for example, by hosting theory in a version-controlled remote repository (such as git), and archiving that repository on Zenodo for long-term storage.
The resulting resource will then have an identifier (DOI) which allows the theory to be accessed using a standardized communications protocol (download via `https` or `git`).
Secondly (A2), theory metadata should be accessible, even when the theory is no longer available,
which is also achieved via long-term storage (e.g., on Zenodo).
Git remote repositories allow for access control,
and Zenodo allows for access control of individual files/resources.
An unavailable theory typically refers to a theory that was abandoned in favor of a better or more general theory (such as the phlogiston theory, which was superseded by thermodynamics).
In general, it makes sense to keep outdated theories, in order to be able to track the genesis of theories over time, yet, we require the availability of meta data as a minimum requirement.

@@

A third impediment arises when theories have a "dependency on the author" (DOA). <!-- AB citation needed?! Who came up with this term? -->
DOA occurs when a theory cannot be understood by independent scholars,
thus requiring the original author for interpretation and clarification.
We have heard DOA referred to apocryphally as the "ask Leon" phenomenon,
as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
DOA relates to the discourse on "Great Man Theorizing" [@guestWhatMakesGood2024] because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
which violates checks and balances of scientific research.
DOA also renders theories immune to refutation,
because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target  [@szollosiArrestedTheoryDevelopment2021].
The fact that DOA is inherently problematic is illustrated by cases where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
This demonstrates that original authors are not the ultimate authority on their theories.
DOA thus unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.

@@

Interoperability pertains to the property of information artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
The original interoperability principles can be rephrased somewhat to apply to theory.
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or multi-interpretable bitmap image falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated as 
as a type of data that is human- and machine-readable with as few interpretative steps as possible,
as previously recommended [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae, <!-- AB: US readers may expect formulas here ? not sure-->
and statements expressed in some axiomatic system.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very interoperable,
the same image represented in the DOT language <!--AB citation needed; also Aaron has his own Julia package for graph language specs - not sure whether Aaron wants to refer to this?--> for representing graphs does meet this ideal.

@@

If take cumulative knowledge acquisition to be a goal of scientific research, then Reusability is the ultimate purpose of making theory FAIR.
Applied to FAIR theory, reusability requires that  theory and its associated metadata are richly described with a plurality of accurate and relevant attributes (R1) with a clear and accessible license for reuse (R1.1), detailed provenance (R1.2), and (meta)data which meets domain-relevant community standards (R1.3).
As we will argue below, the most appropriate license for theory reuse is likely to be CC0 (no rights reserved), <!-- AB : maybe add a caveat that there may be country-specific legislation, company property rights, etc. and that this is of course no legal advice-->
although this should be combined with a culture of comprehensive (theory) citation to meet other open science requirements [REF TOP guidelines].
Criterion R1.2 is met by version control with Git and archival on Zenodo.
Domain-relevant community standards, to a large extent, remain to be established - and this paper is the first step towards further work in that area.
